Download and plot data for the holdings at NCBI GEO
========================================================

Load all the required packages
```{r}
library(devtools)
install_github("ropensci/plotly")
library(plotly)
library(XML)
```

Setup the plotly API. You will need to have a Plotly account and need your username and API key
```{r}
py <- plotly(username=username, key=key)
```

First, download the historical data from NCBI GEO and get it into a format for plotting:
```{r}
# GEO history data
history <- "http://www.ncbi.nlm.nih.gov/geo/summary/summary.cgi?type=history"
geoDat <- readHTMLTable(doc=history)
geoDat <- as.data.frame(geoDat)

# get the number of samples for each period
samples <- as.character(geoDat$NULL.Samples)

# Remove the comma from the numbers
samples <- gsub(samples, pattern=",", replacement="")
samples <- as.numeric(samples)

# get the series (datasets) for each period and remove the comma from the numbers
series <- as.character(geoDat$NULL.Series)
series <- gsub(series, pattern=",", replacement="")
series <- as.numeric(series)

# get the year and quarter and combine into number reflecting time period
year <- as.numeric(as.character(geoDat$NULL.Year))
quarter <- as.numeric(as.character(geoDat$NULL.Quarter))
date <- year + (quarter/4)
geoSamples <- data.frame(Samples = samples, Year = date, Datasets = series)
```

Plot the data using plotly. Remember to set the Plotly username and key:
```{r}
# Geo data holdings over time
trace1 <- list(
  x = date, 
  y = samples, 
  name = "Samples", 
  type = "bar"
)
trace2 <- list(
  x = date, 
  y = series, 
  name = "Datasets", 
  yaxis = "y2", 
  type = "scatter"
)
data <- list(trace1, trace2)
layout <- list(
  title = "NCBI GEO data holdings 2000-2014", 
  yaxis = list(title = "Number of samples"), 
  font = list(size = 16),
  legend = list(
    x = 0.85, 
    y = 0.9
  ),
  yaxis2 = list(
    title = "Number of data sets", 
    titlefont = list(color = "rgb(148, 103, 189)"), 
    tickfont = list(color = "rgb(148, 103, 189)"), 
    overlaying = "y", 
    side = "right"
  )
  
)

response <- py$plotly(data, kwargs=list(layout=layout, filename="geo-history", fileopt="overwrite"))
url <- response$url
```

Get the breakdown of taxanomic groups:
```{r}
# GEO taxa data
taxa <- "http://www.ncbi.nlm.nih.gov/geo/summary/summary.cgi?type=tax"
geoTaxa <- readHTMLTable(doc=taxa)
geoTaxa <- as.data.frame(geoTaxa)

# Extract the sample numbers for each taxa and remove the commas from the numbers
sTaxa <- as.character(geoTaxa[ ,4])
sTaxa <- gsub(sTaxa, pattern=",", replacement="")
sTaxa <- as.numeric(sTaxa)

# Extract the number of datasets for each taxa
dTaxa <- as.character(geoTaxa[ ,2])
dTaxa <- gsub(dTaxa, pattern=",", replacement="")
dTaxa <- as.numeric(dTaxa)

# Extract the taxa names
nTaxa <- geoTaxa[ ,1]

# Form the data frame for plotting and sort smallest to largest and add colour
geoTaxa <- data.frame(Taxa = nTaxa, Samples = sTaxa, Datasets = dTaxa)
geoTaxa <- geoTaxa[order(geoTaxa$Samples), ]


# Set the Vitamin C colour palette
vitC <- c("#1695A3", "#BEDB39", "#FFE11A", "#1F8A70", "#FD7400", "#004358")
geoTaxa$col <- c(rep(x="#CCCC9F", times=(nrow(geoTaxa) - length(vitC))), vitC)

taxaSamples <- function(taxa){
        taxa <- list(
  x = "Species", 
  y = c(geoTaxa[geoTaxa$Taxa == as.character(taxa), 2]),
  name = as.character(taxa),
  marker = list(color = geoTaxa[geoTaxa$Taxa == as.character(taxa), 4]),
  type = "bar")
}

traces <- lapply(X=geoTaxa$Taxa, FUN=taxaSamples)

layout <- list(barmode = "stack", autosize=FALSE, width = 500, height = 550, title = "Species composition of NCBI GEO data holdings", yaxis = list(title = "Number of samples"), font = list(size = 16))

response <- py$plotly(traces, kwargs=list(layout=layout, filename="geo-taxa", fileopt="overwrite"))

url <- response$url
```

Calculate number of samples per dataset
```{r}
geoSamples <- geoSamples[1:56, ]
geoSamples$setSize <- geoSamples$Samples/geoSamples$Datasets
barplot(geoSamples$setSize)

abline()


# Geo data holdings over time
py <- plotly(username=username, key=key)

trace1 <- list(
  x = geoSamples$Year, 
  y = geoSamples$setSize, 
  name = "Average dataset size (samples)", 
  type = "bar"
)

data <- list(trace1)
layout <- list(
  title = "NCBI GEO dataset sizes 2000-2014", 
  yaxis = list(title = "Number of samples per dataset"), 
  font = list(size = 16),
  legend = list(
    x = 0.85, 
    y = 0.9
  )
)

response <- py$plotly(data, kwargs=list(layout=layout, filename="geo-dataset-size", fileopt="overwrite"))
url <- response$url
```


GEO series type data
```{r}
# GEO type data
type <- "http://www.ncbi.nlm.nih.gov/geo/summary/summary.cgi?type=series"
geoType <- readHTMLTable(doc=type)
geoType <- as.data.frame(geoType)

# Extract the sample numbers for each taxa and remove the commas from the numbers
sType <- as.character(geoType[ ,2])
sType <- gsub(sType, pattern=",", replacement="")
sType <- as.numeric(sType)

# Extract the taxa names
nType <- geoType[ ,1]

# Form the data frame for plotting and sort
geoType <- data.frame(SeriesType = nType, Samples = sType)
geoType <- geoType[order(geoType$Samples, decreasing=TRUE), ]

geoTypeTop <- geoType[1:7, ]
geoTypeOther <- geoType[8:nrow(geoType), ]
geoTypeTop[8, 1] <- "Other"; geoTypeTop[8, 2] <- sum(geoTypeOther$Samples)
geoTypeTop <- geoTypeTop[8:1, ]

typeSamples <- function(x){
        taxa <- list(
  x = "Series Type", 
  y = c(geoType[geoType$SeriesType == as.character(x), 2]),
  name = as.character(x),
  marker = list(color = geoType[geoType$SeriesType == as.character(x), 4]),
  type = "bar")
}

traces <- lapply(X=geoTypeTop$SeriesType, FUN=typeSamples)

layout <- list(barmode = "stack", autosize=FALSE, width = 500, height = 550, title = "Number of samples for each GEO series type", yaxis = list(title = "Number of datasets"), font = list(size = 16))

response <- py$plotly(traces, kwargs=list(layout=layout, filename="geo-series-type", fileopt="overwrite"))

url <- response$url


```

